Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-06 ms.tex
6.6

Low-mass AGN in the local universe ($M_{\bullet} \lesssim 3\times
10^{7}M_{\odot}$) are observed to be growing on a timescale comparable
to the age of the universe, while the most massive SMBHs with
$M_{\bullet} \gtrsim 10^{9}M_{\odot}$ possess local growth times which
are more than 2 orders of magnitude longer (\citealt{Heckman+04};
\citealt{Kauffmann&Heckman09}).

The bottom panel of Figure~\ref{fig:bh_growth} shows the SMBH growth
time, $t_{\rm grow} = M_{\bullet}/\dot{M}$, for each accretion rate
shown in the top panel.  For black holes accreting steadily at the rate set by the
stellar wind heating associated with the average star formation
history of their host galaxies, we see that $t_{\rm grow}$ exceeds the
Hubble time by 2$-$3 orders of magnitude across all black hole masses.
Steady accretion therefore cannot explain the growth of low mass black
holes, a fact which is not surprising given that approximately half of
this growth occurs in AGN radiating within 10 per cent of their
Eddington rate (\citealt{Heckman+04}).  Such high accretion rates
likely instead require an source of gas external to the nuclear
region, triggered either by galaxy mergers associated with the
hierarchical growth of structure or thermal instabilities on larger,
galactic scales (e.g.~\citealt{Ciotti+10}).  In support of this,
\citet{Kauffmann&Heckman09} find that the growth of low mass black
holes is dominated by young stellar bulges which have a log normal
distribution of Eddington ratios peaked at $\Mdot/\MdotEdd\sim
0.01$. That this distribution is insensitive to galaxy properties
suggests that it is set by SMBH feedback instead of the nuclear gas
supply.

That the growth time associated with thermally-unstable accretion
(yellow line) exceeds the Hubble time across all SMBH masses
highlights the fact that significant black hole growth in the local
universe cannot result from thermally-stable steady accretion of gas
lost from the surrounding stellar population studied in this paper.
Gas blow-out by SN Ia cannot prevent the growth of low-mass black
holes, as indicated by the low growth times $\ll t_{\rm h}$ allowed by
Ia heating (green lines), although Ia heating could play in principle
a role in capping the growth rate of $\gtrsim 10^{8}M_{\odot}$ black
holes, again depending on the efficiency of Ia heating and whether the
galaxy is a core or cusp.

@@

Our results also have implications for the environments encountered by
relativistic jets from TDEs.  For low-mass SMBHs $M_{\bullet} \lesssim
10^{7}M_{\odot}$, characterizing those responsible for {\it Swift}
J1644+57 (\citealt{Bloom+11}), we predict gas densities of a $n
\sim 0.3-3$ cm$^{-3}$ on radial scales of 0.3 pc for wind heating
rates within the physically expected range $v_{\rm w} \sim 500-1000$
km s$^{-1}$ (Fig.~\ref{fig:profiles}).  These values are within the
range estimated by modeling the radio afterglow of {\it Swift}
J1644+57 (\citealt{Metzger+12}, \citealt{BergerZauderer+:2012a}), as shown in Fig.~\ref{fig:bh_growth}.

\citet{BergerZauderer+:2012a} infer a flattening of the gas density
profile of the host of J1644+57 on radial scales $r \gtrsim 0.4$ pc (their Fig.~6) which looks
qualitatively similar to the shape density profile we predict for core
galaxies (Fig.~\ref{fig:profiles}, solid line).  To obtain such a flattening on the inferred radial scales would require a black hole of $\Mbh\sim 10^7 \Msun$.  However,  this is inconsistent with the break in the gas density seen at $\sim 0.6$ pc by \citet{BergerZauderer+:2012a}, which should correspond to the break radius of the {\it stellar density} profile, thus implying an unrealistically small SMBH mass of $\Mbh\simeq 100\Msun$ according to the mean $r_b-M_{\bullet}$ relationship \citep{LauerFaber+:2007a}.  Alternatively, the observed density break in J1644+57 could result from the outer edge of a sub-parsec nuclear star cluster (e.g., \citealt{Carson+15}).  Such compact star clusters could greatly enhance the rate of TDEs (e.g.~\citealt{Stone&Metzger15}), making this association uncoincidental.

@@

\subsection{Regulation by Kinetic Outflow Feedback}
\label{sec:kinetic}

Our analysis of black hole feedback has focused on Compton heating
instead of kinetic feedback, since the efficacy of radiative feedback
is relatively straightforward to calculate given the properties of the
accretion flow.  However, if kinetic feedback instead dominates, this
could play an important role in regulating the accretion.  Our simple
parameterization of kinetic feedback (eq.~[\ref{eq:vjet}]) predicts an
effectic wind velocity which increases with SMBH mass as
$v_{w}^{\bullet} \propto M_{\bullet}^{0.33(0.28)}$ for core(cusp)
galaxies, respectively.  Coupled with the dependence of the SMBH
accretion rate on the wind heating, $\dot{M}/\dot{M}_{\rm edd} \propto
M_{\bullet}^{0.76(0.48)}v_{w}^{-3.8(-2.4)}$ (Fig.~\ref{fig:mdot_mass};
eq.~[\ref{eq:eddr_analytic}]), kinetic feedback of the form we have
adopted leads to an Eddington ratio dependence on SMBH mass of
$\dot{M}/\dot{M}_{\rm edd} \propto M_{\bullet}^{-0.5(-0.2)}$.

This prediction is in better with the observed dependence of $\langle
L_{X}/L_{\rm edd} \rangle$ in elliptical galaxies
(Fig.~\ref{fig:miller}, green line), hinting that accretion regulation
by kinetic feedback may serve as an alternative explanation for X-ray
luminosities of elliptical galaxies.  However, our assumption that
kinetic outflows heat the gas uniformly in volume is somewhat
arbitrary and would need to more rigorously justified by numerical
studies of how jets or disk outflows couple energy to their gaseous
environment.  It is furthermore unclear whether our steady-state model
can be applied at all, because black hole feedback depends on the
central accretion rate, which (unlike stellar feedback) does not
respond instantaneously to changes in the thermodynamic properties of
the gas on larger scales.

>>>>

Low-mass AGN in the local universe with $M_{\bullet} \lesssim 3\times
10^{7}M_{\odot}$ are observed to be growing on a timescale comparable
to the age of the universe, while the most massive SMBHs with
$M_{\bullet} \gtrsim 10^{9}M_{\odot}$ possess local growth times which
are more than 2 orders of magnitude longer (\citealt{Heckman+04};
\citealt{Kauffmann&Heckman09}).

The bottom panel of Figure~\ref{fig:bh_growth} shows the SMBH growth
time, $t_{\rm grow} \equiv M_{\bullet}/\dot{M}$, for each accretion rate
shown in the top panel.  For SMBHs which accrete steadily at the rate set by the
stellar wind heating due to the average star formation
history of their host galaxies, we see that $t_{\rm grow}$ exceeds the
Hubble time by 2$-$3 orders of magnitude across all $M_{\bullet}$.
Steady accretion therefore cannot explain the growth of low mass black
holes, a fact which is not surprising given that approximately half of
this growth occurs in AGN radiating within 10 per cent of their
Eddington rate (\citealt{Heckman+04}).  Such high accretion rates
likely instead require an source of gas external to the nuclear
region, triggered either by galaxy mergers associated with the
hierarchical growth of structure or thermal instabilities on larger,
galactic scales (e.g.~\citealt{Ciotti+10}).  In support of this,
\citet{Kauffmann&Heckman09} find that the growth of low mass black
holes is dominated by young stellar bulges which have a log normal
distribution of Eddington ratios peaked at $\Mdot/\dot{M}_{\rm edd} \sim
0.01$. That this distribution is insensitive to galaxy properties
suggests that it is set by SMBH feedback instead of the nuclear gas
supply.

That the growth time associated with thermally-unstable accretion
(yellow line in Fig.~\ref{fig:bh_growth}) exceeds the Hubble time across all SMBH masses
highlights the fact that significant black hole growth in the local
universe cannot result from thermally-stable steady accretion of gas
lost from the surrounding stellar population studied in this paper.
Gas blow-out by SN Ia cannot prevent the growth of low-mass black
holes, as indicated by the low growth times $\ll t_{\rm h}$ allowed by
Ia heating (green lines), although Ia heating could play in principle
a role in capping the growth rate of $\gtrsim 10^{8}M_{\odot}$ black
holes, again depending on the efficiency of Ia heating and whether the
galaxy is a core or cusp.

@@

Our results also have implications for the environments encountered by
relativistic jets from TDEs.  For low-mass SMBHs $M_{\bullet} \lesssim
10^{7}M_{\odot}$, such as that responsible for powering the transient {\it Swift}
J1644+57 (\citealt{Bloom+11}), we predict gas densities of a $n
\sim 0.3-3$ cm$^{-3}$ on radial scales of 0.3 pc for wind heating
rates within the physically expected range $v_{\rm w} \sim 500-1000$
km s$^{-1}$ (Fig.~\ref{fig:profiles}).  These values are within the
range estimated by modeling the radio afterglow of {\it Swift}
J1644+57 (\citealt{Metzger+12}, \citealt{BergerZauderer+:2012a}), as shown in Fig.~\ref{fig:bh_growth}.

\citet{BergerZauderer+:2012a} infer a flattening of the gas density
profile of the host of J1644+57 on radial scales $r \gtrsim 0.4$ pc (their Fig.~6) which looks
qualitatively similar to the shape density profile we predict for core
galaxies (Fig.~\ref{fig:profiles}, solid line).  To obtain such a flattening on the inferred radial scales would require a black hole of $\Mbh\sim 10^7 \Msun$.  However,  this is inconsistent with the break in the gas density seen at $\sim 0.6$ pc by \citet{BergerZauderer+:2012a}, which should correspond to the break radius of the {\it stellar density} profile, thus implying an unrealistically small SMBH mass of $\Mbh\simeq 100\Msun$ according to the mean $r_b-M_{\bullet}$ relationship \citep{LauerFaber+:2007a}.  Alternatively, the observed density break in J1644+57 could result from the outer edge of a sub-parsec nuclear star cluster (e.g., \citealt{Carson+15}).  The high stellar densities of such a compact star cluster could greatly enhance the TDE rate (e.g.~\citealt{Stone&Metzger15}), possibly making this association uncoincidental.

@@

\subsection{Regulation by SMBH Kinetic Feedback}
\label{sec:kinetic}

Our analysis of SMBH feedback has focused on Compton heating
instead of kinetic feedback, since the effects of radiative feedback
are relatively straightforward to calculate from the properties of the
accretion flow.  However, it is possible kinetic feedback from SMBH outflows could play an equal or greater role in regulating accretion.  Our simple
parameterization of kinetic feedback (eq.~[\ref{eq:vjet}]) assuming heating which is volumetrically uniform predicts an
effectic wind velocity which increases with SMBH mass as
$v_{w}^{\bullet} \propto M_{\bullet}^{0.33(0.28)}$ for core(cusp)
galaxies, respectively.  Coupled with the dependence of the SMBH
accretion rate on the wind heating, $\dot{M}/\dot{M}_{\rm edd} \propto
M_{\bullet}^{0.76(0.48)}v_{w}^{-3.8(-2.4)}$ (Fig.~\ref{fig:mdot_mass};
eq.~[\ref{eq:eddr_analytic}]), a dominant source of kinetic feedback of the form we have
adopted leads to an Eddington ratio dependence on SMBH mass of
$\dot{M}/\dot{M}_{\rm edd} \propto M_{\bullet}^{-0.5(-0.2)}$.

This prediction is also consistent with the observed dependence of $\langle
L_{X}/L_{\rm edd} \rangle$ in elliptical galaxies
(Fig.~\ref{fig:miller}, green line), hinting that accretion regulation
by kinetic feedback may serve as an alternative explanation for X-ray
luminosities of elliptical galaxies.  However, our assumption that
kinetic outflows heat the gas uniformly in volume is somewhat
arbitrary and would need to more rigorously justified by numerical
studies of how jets or disk outflows couple energy to their gaseous
environment.  It is furthermore unclear whether our steady-state model
can be applied at all, because black hole feedback depends on the
central accretion rate, which (unlike stellar feedback) does not
respond instantaneously to changes in the thermodynamic properties of
the gas on larger scales.
